---
title: R code for the development of short-term predictions of E.coli and biotoxins
  in shellfish
author: "Wiebke Schmidt, Jez Smith and Rob Thomas"
output:
  word_document: default
---

# README
Please find below the R code for the development of short-term predictions of E.coli and biotoxins in shellfish, as used within the publication: Schmidt et al., 2018. "A generic approach for the development of short-term predictions of Escherichia coli and biotoxins in shellfish". Aquaculture Environment Interactions 10:173-185, https://doi.org/10.3354/aei00265  

This work was carried out as part of the UK Biotechnology and Biological Science Research Council (BBRSC) and UK National Environmental Research Council
(NERC) funded 'ShellEye' project (BB/M026698/1).   

Due to the confidential nature of some of the research materials supporting this publication, not all of the data can be made accessible to other researchers. Part of the code is based on the work by Jez Smith & Rob Thomas (www.eco-explore.co.uk) and you are kindly asked if you use this R code, please cite as below:  

authors: Schmidt, W., Smith, J. and Thomas, R.  
orcid: https://orcid.org/0000-0003-1035-6054  
title: "R code for the development of short-term predictions of E.coli and biotoxins in shellfish"  
version: 1.0  
doi: 10.5281/zenodo.2598466   
date-released: 2019-03-19  

# R code
Please not the R code & variable names have been defined in general terms, please adjust to your own dataset. Additioanlly, please note that the "eval = FALSE" parameter was added to the code chunk, please change to "eval = TRUE"" to run the code.  

## 1. Load libraries
If not installed, use install.packages("")
```{r load libraries, eval=FALSE}
# Load libraries, if not installed, use install.packages("")
library(leaps)
library(bestglm)
library(arm) 
library(MuMIn)
library(car)
library(nlme)
library(mgcv) 
library(lattice)
```

## Generalized linear model development
```{r GLM development, eval=FALSE}
# Load data and define variables, such as rainfall = rain, etc.

# Create data frame of environmental variables and E.coli/ biotoxin as y
Xybt1<-data.frame(cbind (rain, river, sst, windspeed, winddirection, solarradiation, y))

# Method 'CV', default cross-validation method is delete-d with 1000 replications
model1<-bestglm(Xybt1, family=gaussian, IC="CV", t=10,method='exhaustive') 
print.bestglm(model1)
summary.bestglm(model1)
# RMSE for model 1
MSE1<-model1$Subsets[,"CV"]
round(sqrt(mean(MSE1)),2)

# method 'HTF' based on Hastie et al., 2009
model2<-bestglm(Xybt1,family=gaussian,IC="CV", CVArgs = list(Method="HTF", K=10, REP=50),method='exhaustive')                                
print.bestglm(model2)
summary.bestglm(model2)
# RMSE for model 2
MSE2<-model2$Subsets[,"CV"]
round(sqrt(mean(MSE2)),2)

# method 'DH' based on Davison and Hinkley (1997, Algorithm 6.5, p.295)
model3<-bestglm(Xybt1,family=gaussian,IC="CV", CVArgs = list(Method="DH", K=2, REP=100), method='exhaustive')
print.bestglm(model3)
summary.bestglm(model3)
# RMSE for model 3
MSE3b<-model3$Subsets[,"CV"]
round(sqrt(mean(MSE3)),2)

# method Leave-one-out Cross-Validation
model4<-bestglm(Xybt1,family=gaussian,IC="LOOCV")
print.bestglm(model4)
summary.bestglm(model4)
# RMSE for model 4
MSE4b<-model4$Subsets[,"LOOCV"]
round(sqrt(mean(MSE4)),2)
```

## Averaged GLM development
```{r average GLM development, eval=FALSE}
# Build a "global model" incorporating all main effects and/or interactions of interest

model5<-glm(y ~ rain + river + sst + windspeed + winddirection + solarradiation,
            family = Gamma (link = log))

summary(model5)
summary.lm(model5)
plot(model5) 
AIC(model5)

# Standardise the global model's parameters
stdz.model5 <- standardize(model5, standardize.y = FALSE) 
summary(stdz.model5)

# Create a set of candidate models 
options(na.action = "na.fail")
model.set <- dredge(stdz.model5)
model.set

# Select the top models, using delta Akaike's Information Criterion < 2 as decision metric  
top.models <- get.models(model.set, subset = delta <2)
top.models

# Obtain parameter estimates averaged across the top.models set
averaged.model <- model.avg(top.models)  
summary(averaged.model)
# Check variance inflation factors 
vif(model5) 
```

## Generalized additive model development
```{r GAM development, eval=FALSE}
# Create model
model6<-gam(y ~ s(rain, k=4) + 
                s(river, k=4)+ 
                s(sst, k=4)  + 
                s(windspeed, k=4) + 
                s(winddirection, k=4) +
                s(solarradiation, k=4),family = Gamma (link = "log")) 

# Check model & EDF & adjust model, if necessary
summary (model6) 
gam.check(model6)
plot(model6, shade=T, scale=0) 
# compare adjusted GAM models using Akaike Information Criterion (AIC)
AIC()
```

## Evaluation & Prediction of developed models
```{r Evaluation & Prediction of models, eval=FALSE}
# 1. Re-run choosen best models (glm, average glm and gam)
# 2. Make Predictions using independent dataset 
glmPredictions<-predict(bestglm, newdata=newdata, type="response", se.fit = T)
avglmPredictions<-predict(newavmodel, newdata=newdata, type="response", se.fit = T) 
gamPredictions<-predict(bestgam, newdata=newdata, type="response", se.fit = T)
# Create a table of new predictions and save as csv.file
Predictions<-cbind(newdata,glmPredictions,avglmPredictions, gamPredictions) 
write.csv(Predictions, "prediction-outcome.csv", row.names=F)
# 3. Use RMSE and Bias to assess performance of model prediction in comparions to actual observations (Ecoli/ Biotoxin)
```
